---
title: "Patel_D_10"
output: html_notebook
---
Cereal Data set

we will first work with the cereal data set and perform various tasks on it

```{r}
require(readxl)
# Import the excel file into a data frame 
cereal_data <- read_xls("cereal_data__for_slide_deck_.xls", sheet = 2)

cereal_data
```
1. Using the Cereal Data set, determine the mean, median, range, mode, and standard deviation of the carbs and fiber variables
```{r}
library(dplyr)

carb_mean <- mean(cereal_data$Carbs)
fiber_mean <- mean(cereal_data$Fiber)
sprintf("The mean for carbs: %g", carb_mean)
sprintf("The mean for fiber: %g", fiber_mean)

carb_median <- median(cereal_data$Carbs)
fiber_median <- median(cereal_data$Fiber)
sprintf("The median for carbs: %g", carb_median)
sprintf("The median for fiber: %g", fiber_median)

carb_range <- range(cereal_data$Carbs)
fiber_range <- range(cereal_data$Fiber)
sprintf("The range for carbs: %s - %s", carb_range[1], carb_range[2])
sprintf("The range for fiber: %s - %s", fiber_range[1], fiber_range[2])

carb_mode <- as.numeric(names(table(cereal_data$Carbs))[1])
fiber_mode <- as.numeric(names(table(cereal_data$Fiber))[1])
sprintf("The mode for carbs: %g", carb_mode)
sprintf("The mode for fiber: %g", fiber_mode)

carb_std <- sd(cereal_data$Carbs)
fiber_std <- sd(cereal_data$Fiber)
sprintf("The standard deviation for carbs: %g", carb_std)
sprintf("The standard deviation for fiber: %g", fiber_std)

```

2. Histograms for calories and sugar variables
```{r}
library(ggplot2)
ggplot(data = cereal_data, aes(cereal_data$Calories)) +  geom_histogram() +
  labs(title="Histogram for Calories") +
  labs(x="Calories", y="Cereals Count") 
```
```{r}
library(ggplot2)
ggplot(data = cereal_data, aes(cereal_data$Sugars)) +  geom_histogram() +
  labs(title="Histogram for Sugars") +
  labs(x="Calories", y="Cereals Count") 
```

3. Scatterplots of the Carbs and Fiber variables
```{r}
ggplot(data = cereal_data, aes(x = Carbs, y = Fiber)) + geom_point() +
  labs(title="Scatterplot Carbs vs Fiber")
```

4. Pearson Moment and the Spearman Rank correlation coefficients for the calories and fiber variables
```{r}
# The Pearson Moment Formula:
# z(k) = k - mean(k) / std(k)
# Pearson Moment = sum(1, n, z(x) * z(y)) / n - 1
cal_fib_moment <- cor(cereal_data$Calories, cereal_data$Fiber)
cal_fib_moment

```
There is a negative correlation between the amount of calories and fiber in data. As calories rise fiber falls.

```{r}
# Spearman Rank Correlation
cal_fib_rho <- cor(cereal_data$Calories, cereal_data$Fiber, method = "spearman")
cal_fib_rho

```
The speaman coefficient also shows a negative correlation. Calories up = fiber down

##########################################################################################
Grades Data Set

```{r}
require(openxlsx)
# Import the data for fall and spring into seperate data frames
fall_grades_data <- read.xlsx("workshop_data_set__grades_2018.xlsx", sheet = 1, startRow = 2)
spring_grades_data <- read.xlsx("workshop_data_set__grades_2018.xlsx", sheet = 2, startRow = 3)

head(fall_grades_data)
```
```{r}
print(head(spring_grades_data))
```

1. Using the Grades Data set, determine the mean, median, range, mode, and standard deviation of the Score variable for Fall and Spring.
```{r}

fall_mean <- mean(fall_grades_data$Score)
spring_mean <- mean(spring_grades_data$Score)
sprintf("The mean for fall: %g", fall_mean)
sprintf("The mean for spring: %g", spring_mean)

fall_median <- median(fall_grades_data$Score)
spring_median <- median(spring_grades_data$Score)
sprintf("The median for fall: %g", fall_median)
sprintf("The median for spring: %g", spring_median)

fall_range <- range(fall_grades_data$Score)
spring_range <- range(spring_grades_data$Score)
sprintf("The range for fall: %s - %s", fall_range[1], fall_range[2])
sprintf("The range for fall: %s - %s", spring_range[1], spring_range[2])

fall_mode <- as.numeric(names(table(fall_grades_data$Score))[1])
spring_mode <- as.numeric(names(table(spring_grades_data$Score))[1])
sprintf("The mode for fall: %g", fall_mode)
sprintf("The mode for spring: %g", spring_mode)

fall_std <- sd(fall_grades_data$Score)
spring_std <- sd(spring_grades_data$Score)
sprintf("The standard deviation for fall: %g", fall_std)
sprintf("The standard deviation for spring: %g", spring_std)
```
2. Pearson Moment and the Spearman Rank correlation coefficients for the fall score and excel quiz variables
```{r}
# The Pearson Moment Formula:
# z(k) = k - mean(k) / std(k)
# Pearson Moment = sum(1, n, z(x) * z(y)) / n - 1
score_quiz_moment <- cor.test(fall_grades_data$Score, fall_grades_data$Excel.Quiz)
score_quiz_moment

```
There is a positive correlation between the amount of calories and fiber in data. As scores rise quiz grades rise. It is statistically significant.

```{r}
# Spearman Rank Correlation
score_quiz_rho <- cor.test(fall_grades_data$Score, fall_grades_data$Excel.Quiz, method = "spearman")
score_quiz_rho

```
The speaman coefficient also shows a positive correlation. Score up = Excel Quiz up, the pvalue is small and so its is
statistically significant

3. Scatter plot using the Fall score and the Excel Quiz variable
```{r}
ggplot(data = fall_grades_data, aes(x = Score, y = Excel.Quiz)) + geom_point() +
  labs(title="Scatterplot Score vs Excel Quiz")
```

```{r}
binning_frame <- data.frame(bins=integer(1000))

for(i in 1:1000){
binning_frame[i, "bins"] <- mean(sample(fall_grades_data$Score,round(length(fall_grades_data) * .2)))
}

binning_frame

bins <- cut(binning_frame[, "bins"], 100, include.lowest = TRUE)

plot(bins)

```
From this plot we can see the distribution is skewed left and most scores are concentrated in the 93-96 range.

